---
output:
  html_document:
    df_print: paged
    # toc: true
    # toc_float: true
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

```{css echo=FALSE}
.excc{
background-color: lightblue;
}
```

\newpage 

\tableofcontents

\newpage 



\newpage 

## Course 1 - Introduction to R  {.tabset}


### Install and navigate R Studio.

R is a free software environment for statistical analysis and data science. It is a collaborative effort started by Robert Gentleman and Ross Ihaka in New Zealand, but now made up of hundreds of people. R is made freely available by the Comprehensive R archive Network (CRAN), in the UK it can be downloaded through the University of Bristol [here](https://www.stats.bris.ac.uk/R/). 

Downloading R gives you the basic software environment, but an incredibly popular add-on called 'RStudio' is required to follow this course. You should download the free 'RStudio Desktop Open Source Licence' version for the laptop you will be attending the course with from [RStudio.com](https://www.rstudio.com/products/rstudio/#Desktop).

If you have time before the course, it would be hugely beneficial to get familiar with RStudio.

**Objectives:**

> Download R from <https://www.stats.bris.ac.uk/R/>.  
> Download RStudio from <https://www.rstudio.com/products/rstudio/#Desktop>.

<br><br>

### __Basics I__

Our course starts in the R console, which those of you who are familiar with R but not RStudio will recognise. We will enter commands as input into the console, and receive output from the console. We will start with some simple basic operations, for which R is clearly very excessive.

#### Basic operations
Entering 1+1, we get the output [1] 2. The output is 2, but the [1] lets us know that the number 2 is the first number of output. If we had an output that was particularly large (e.g. 100 seperate numbers) then r may let us know that the first row displayed starts with the first value [1] and the second row starts with the [x]th value. 

```{r,echo=TRUE}

rm(list = ls())

## add 1 to 1. 
1 + 1

## divide 12 by 4
12/4

## times 3 by 7
3*7

## 10 to the power 3
10^3

## root isn't a basic operation so we will look at later.
```
<br><br>

#### Objects
It is possible to create objects, an object can take a number of forms (e.g. a number, a vector of numbers, a matrix, a data-frame). We can then use these objects going forward rather than the values directly. Operations can be applied to these objects, and objects can be over-written. R is basically just objects and functions.

```{r,echo=TRUE}

## objects basics
x <-  3
y <-  5
x + y
 
x <-  4
x + y
  
z <- x + y
z

```
<br><br>

#### Overwriting / Manipulating Objects

```{r,echo=TRUE}

 
a <- 10
a

a <- a + 1 
a 

```

<br><br>

#### *Exercise 1*

```{excc, class.source="excc"}
  1. Set d equal to 10.
  2. divide d by 5
  3. multiply d by 8
  4. add 8 to d 
  5. what is d?
```

<details><summary>Show solution</summary><p>
```
  1. d = 10
  2. d = d / 5
  3. d = d * 8
  4. d = d + 8
  5. d
```
</p></details>

<br><br>

### __Basics II__

#### Evaluations
We can perform evaluations, which provide a true or false answer. For example the input 4>2 returns "FALSE". 

It can be very useful in cases where an outcome is binary (e.g. an individual dies or remains alive). Or where we want to change a continous variable to a binary.

```{r,echo=TRUE}

## simple evaluations
## 4 is greater than 2
4 > 2
## 4 is greater than 5
4 > 5
## 4 is equal to 3, note double == for an evaluation
4 == 3
## 4 is not equal to 3, note != is not equal to.
4 != 3
## the character x is equal to the character x.
"dog" == "dog"
"dog" == "cat"

## the output from an evaluation can be stored as an object, x. This object can be subject to operations & manipulations.
b <- 4<2
b
```
<br><br>

#### Different object classes and types
Objects don't have to take a single value. For example a single object may be the heights of each child in a group of children (in the example below a small class of 4).

We have been working with single values, which are vectors of 1. To illustrate the different classes we are going to create some vectors.

#### Object Classes
Different class include numeric, character, logical, integer & complex (ignore).
```{r,echo=TRUE}

## numeric
height <- c(1.72,1.78,1.65,1.90) ## 1:4
height

## numeric
weight <- c(68,75,55,79)
weight
class(weight)

## character
first_name <- c("Alice","Bob","Harry","Jane")
first_name
#first_name + 1 ## error

## factor
sex <- factor(x = c("F","M","M","F"))
sex

## logical
tall <- height > 1.8
```
<br><br>

#### Operations on different data structures

##### Operations on Vectors

```{r, echo=TRUE}
#Adding:
c(1,2,3) + 1
c(1,2,3) + c(1,2,3)


#multiplication
heightft <- height*3.28

```
<br><br>

#### *Exercise 2*
```{excc, class.source="excc"}
1. Create a vector called 'odds' with the numbers 1,3,5,7,9.
2. Show what class odds is. 

3. Evaluate which numbers in the odd vector are greater than 4. 

4. Create a vector called 'fail' containing 1,3,5,'seven',9.
5. Show what class fail is.

6. Create a vector that gives everyone's weight in pounds (2.2lbs to kg)
````

<details><summary>Show solution</summary><p>
```
  1. odds = c(1,3,5,7,9)
  2. class(odds)
  3. odds > 4
  3. fail = c(1,3,5,'seven',9)
  4. class(fail)
  5. lb_from_weight = weight * 2.2
     lb_from_weight
```
</p></details>

<br><br>


### __Basics III__

#### Basic object Types

There are multiple types of object in R. We can store objects together in a data-frame. In our example data-frame each column is a variable (height, weight, first_name), and each row is an individual. 

Different object types include:

Vector - single variable a 1x1 vector. Vector all elements same data-type. 
Matrix - all same data-type.
Dataframe - columns same data type.
List - anything goes.

```{r,echo=TRUE}

## data frame- columns are variables, rows are observations.
df <- data.frame(height,weight,first_name,sex)
df

## we can select a single variable within the dataframe using the dollar sign.
df$height

## We can add a new variable easily, in this case based on other variables within the dataframe.
df$bmi <- df$weight / df$height^2 
df
```
<br><br>

#### Subsetting
We can subset our data, to reduce it to those we are interested in. This is useful when cleaning our data, and when changing a continuous variable to a categorical.
```{r,echo=TRUE}

## Our data-frame contains the height, weight, first name and bmi of 4 individuals.
df

#To subset a data frame we can use square brackets i.e df[row,column]
#Selecting a column(s)
df$height
df[,"height"]
df[,1]
df[,1:3]
df[,c(1,3)]

#selecting a row(s)
df[1,]

#We might also want to select observations (rows) based on characteristics of the data
#E.g. we might want to only look at the data for people who are taller than 1.75m

#create a logical variable called min_height which contains T/F for each individual being over 175cm.
min_height <- df$height >= 1.75
min_height

## Subset the data to include only those observations (rows) for which height > 175cm (using min_height).
df.at_least_175 <- df[min_height,]
df.at_least_175

#People smaller than 1.75m
## Subset the data to include only those who are not above min-height of 175cm.
smaller = df$height < 1.75
df[smaller,]
df[!min_height,]


```
Note that there are other more advanced methods, which uses pipes and require less code (these are covered in more advanced courses).
<br><br>

#### *Exercise 3*
```{excc, class.source="excc"}
1. Select the 3rd row from the data frame

2. Select the weight variable from the data frame using your prefered method

3. Select alice's data from the data frame 

4. Subset the data frame to show just the data for the females

5. type df[,-1] what does this give
```

<br><br>

#### Exercises

**Exercise 1**
Simple calculations

a) Calculate the following:

5*10
20/3

More complex calculations.
b) Calculate x where a = 20 b = 9, c = 5, d = 1.2
*you are only allowed to type each number once*

$x = 4b + 7c + 3d$

$x = \frac{8b + 4c -12d}{a}$



${12\times(a+b)}\over {x}$

**Exercise 2**
x <- c(10,30,4,52,60,7,8,10,12,15,14,17,19,20,25,30)
a) Which numbers in x are above 8
b) Which numbers are equal to 10.
c) Which numbers are below 8 or above 30.

d) Can you create a matrix with numbers and characters.
        names <- c("Anne","Tom","Jamie","Max","Claire")
        ages <- c(12,16,25,34,28)
        cbind(names,ages)
        What happens if you try to use the ages?


e) Create a dataframe for five individuals (Andrew, Betty, Carl, Diane and Elisa) who are aged (62,80,24,40,56) and have gender (male, female, male,female, female). 

f) Use evaluations and subsetting to find the characteristics of the individual who can claim their free bus pass (age 65+).

g) Create a variable in the dataframe called life expectancy, set this to 83 for females and 80 for males.

h) Create another variable called lyr (life years remaining) which is the number of years to life expectancy for each individual

<br><br>
\newpage


